Polyester RNA-seq simulation study

### 2018-02-16 datasets/RNA-Seq/polyester-replicates-benchmarkFDR.Rmd
updated polyester sims to B=10 reps; deleted old polyester sims

# # set up parallel backend:
library(doParallel)
nCores <- 1
multicoreParam <- MulticoreParam(workers = nCores)
```

@@

```{r}
bd <- initializeBenchDesign() # only needs to be done once
```

Try running `buildBench()` one time.
```{r}
sampleSize <- 5
nDE <- 0

groupPolyester = rep(c(-1,1),each=sampleSize)
modPolyester = model.matrix(~-1 + groupPolyester)

@@

# NULL simulation
truth <- rep(FALSE, nrow(counts)) 
names(truth) <- rownames(counts)

# log2FC are all 0 in null simulation
log2FC = rep(0,nrow(counts))
dat_alt = create_read_numbers(params$mu, params$fit,
                              params$p0, m=dim(counts)[1], 
                              n=sampleSize*2, beta=cbind(log2FC), 
                              mod=modPolyester)

@@

# filter NAs and those with less than 50% expressed 
geneExp  <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)
```

```{r}
head(geneExp)
```

```{r}
# build Benchmark data
sb <- bd %>% buildBench(data=geneExp, parallel=FALSE, BPPARAM=multicoreParam, 
                        truthCols = "qvalue")
```

>>>>

# # set up parallel backend:
library(BiocParallel)
nCores <- 10
multicoreParam <- MulticoreParam(workers = nCores)
```

@@

## Helper functions for simulation

Next, we'll simulate RNA-Seq samples, both with and without the 
addition of simulated DE genes. Here we'll create a function that we can use to 
run one replicate given sample size and number of DE gene settings. This will be
looped over many replications and results averaged over them.

```{r}
#' @param counts is the full count dataset
#' @param paramPolyester is the parameters from running polyester. 
#' Contains information on the mean-variance relationship used to 
#' simulate RNA-Seq counts. 
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
simulateOneSplit <- function(counts, paramPolyester, 
                             nDE, sampleSize, bd, BPPARAM){

  # things needed to simulate with Polyester
  groupPolyester = rep(c(-1,1),each=sampleSize)
  modPolyester = model.matrix(~-1 + groupPolyester)

@@

  # NULL simulation
  if(nDE == 0) {
    truth <- rep(FALSE, nrow(counts)) 
    names(truth) <- rownames(counts)

    # log2FC are all 0 in null simulation
    log2FC = rep(0,nrow(counts))
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

  # DE simulation
  if(nDE > 0){
    # pick random set of nDE genes to add signal to
    DE <- sample(1:nrow(counts), nDE)
    truth <- rep(FALSE, nrow(counts))
    truth[DE] <- TRUE
    names(truth) <- rownames(counts)

    # randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
    log2FC <- rep(0, nrow(counts))
    # log2FC[DE] <- rtruncnorm(nDE, 1, sd=.5) * sample(c(1,-1), size=nDE, replace=TRUE)
    log2FC[DE] <- c(rnorm(nDE, sd = 1.5), rep(0,nrow(counts)-nDE))
    dat_alt = create_read_numbers(paramPolyester$mu, paramPolyester$fit,
                                  paramPolyester$p0, m=dim(counts)[1], 
                                  n=sampleSize*2, beta=cbind(log2FC), 
                                  mod=modPolyester)
  }

@@

  # filter NAs and those with less than 50% expressed 
  geneExp  <- geneExp  %>% na.omit() %>% dplyr::filter(pzero < 0.5)

  # built Benchmark data
  sb <- bd %>% buildBench(data=geneExp, parallel = FALSE, 
                          BPPARAM = BPPARAM, truthCols = "qvalue", 
                          ftCols = "ind_covariate")
  rowData(sb)$log2FC <- geneExp$effect_size

  return(sb)
}

```

We'll also set some parameters that will be common to all simulations. These
include the number of replications, the bench design object, the set of 
methods to exclude in the results plots, and the alpha cutoff level to 
be used when plotting the aggregated Upset results.

```{r}
B <- 10
bd <- initializeBenchDesign() # only needs to be done once
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
ualpha <- 0.05
```

Here's a helper function to return the number of methods with rejections at
a particular alpha level (this helps us determine whether or not to plot the
aggregated upset plot - if there aren't at least 2 methods it will throw an
error, which is a problem for the null simulations).

```{r}
#' @param res standardized metric data.table generated using
#'        standardize_results.
#' @param alpha alpha cutoff
#' @param filterSet which methods to exclude from consideration 
numberMethodsReject <- function(res, alphacutoff, filterSet){
  res <- res %>% 
    filter(is.na(param.alpha) | (param.alpha == alphacutoff)) %>%
    filter(!(blabel %in% filterSet)) %>%
    filter(alpha == alphacutoff) %>%
    filter(performanceMetric == "rejections") %>%
    select(blabel, performanceMetric, value) %>%
    group_by(blabel) %>%
    summarize(mean_value = mean(value)) %>%
    filter(mean_value > 0)
  return(nrow(res))
}

addDefaultMetrics <- function(sb) {
    sb <- addPerformanceMetric(sb, evalMetric = c("TPR", "FPR", "TNR", "FNR", "rejections"),
                               assay = "qvalue")
    return(sb)
}

```

# Sim N5: Null Comparison 5v5

Here we run a null comparison 5 versus 5 RNA-Seq samples. 
This will be done for 100 simulations.

## Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 0

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null5, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))
} else {
  null5 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))
}
```

## Plot average results over replications

Plot results.

```{r}

res5 <- plotsim_standardize(null5, alpha = seq(0.01, 0.20, 0.01))

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res5, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5, met="TNR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

if (numberMethodsReject(res5, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(null5, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, 
          "; skipping upset plot")
}
```

# Sim N10: Null Comparison 10v10

Here we'll repeat the above, but for a null comparison 10 versus 10 samples. 
This will be done for 100 simulations.

## Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 0

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null10, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))
} else {
  null10 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))
}
```

## Plot average results over replications

Plot results.

```{r}
res10 <- plotsim_standardize(null10, alpha = seq(0.01, 0.20, 0.01))

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res10, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10, met="TNR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

if (numberMethodsReject(res10, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(null10, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, "; skipping upset plot")
}

```

# Sim D5: DE Comparison 5v5

Here we'll repeat the above, but for a DE comparison 5 versus 5 samples
and 1500 DE genes are added. This will be done for 100 simulations.

## Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 5
nDE <- 1500

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de5, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))
} else {
  de5 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))
}
```

## Plot average results over replications

Plot results.

```{r}
plotEffectSize <- function(sbl, alpha=0.05, DE=TRUE, covname){

summarize_one_item <- function(object, alpha){
  object <- object[,!( grepl("^ihw", as.character( colData( object )$blabel ) )
                       & colData( object )$param.alpha != alpha )]

  df <- as.data.frame(cbind(as.matrix(rowData(object)), 
                            1*(assays(object)[["qvalue"]]) < 0.05))
  colnames(df)[1] <- "truth" 
  df <- df %>%
    select(truth, covname, bonf, bh,
                  qvalue, contains("ihw"), ashs, "bl-df03", lfdr, 
                  "scott-theoretical", "scott-empirical") %>%
    # rename("scott-theoretical"="scott-t", "scott-empirical"="scott-e") %>%
    gather(method, significant, -truth, -covname) %>%
    group_by(method, truth, significant) %>%
    summarize(mean_effect_size=mean(abs(get(covname))),
              med_effect_size=median(abs(get(covname))),
              min_effect_size=min(abs(get(covname))),
              max_effect_size=max(abs(get(covname))),
              botQ=quantile(abs(get(covname)),0.25),
              topQ=quantile(abs(get(covname)), 0.75)) %>% na.omit()
  df$significant <- as.factor(df$significant)
  levels(df$significant)=c("Not Significant", "Significant")
  df$truth <- as.factor(df$truth)
  levels(df$truth)=c("Not DE", "DE")
  return(df)
}

df <- lapply(sbl, summarize_one_item, alpha=alpha)
df <- bind_rows(df, .id = "rep")
df <- as.tibble(df)

if (DE){
  df <- df %>% filter(truth=="DE")
}else{
  df <- df %>% filter(truth=="Not DE")
}

p <- ggplot(df, aes(x = as.factor(method), y=med_effect_size)) +
  geom_boxplot() +
  facet_grid(significant~., scales="free_y") +
  xlab("Method") +
  ylab(paste0("Median ", covname))

if (DE){
  p <- p + ggtitle(paste0(covname, " of True DE genes by significance at alpha ", 
                       alpha))
}else{
  p <- p + ggtitle(paste0(covname, " of Non-DE genes by significance at alpha ", 
                       alpha))
}

return(p)
}

res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.20, 0.01))

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res5d, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

plotEffectSize(de5, covname="log2FC")
plotEffectSize(de5, covname="log2FC", DE=FALSE)

plotEffectSize(de5, covname="ind_covariate")
plotEffectSize(de5, covname="ind_covariate", DE=FALSE)

if (numberMethodsReject(res5d, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(de5, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, "; skipping upset plot")
}

```

# Sim D10: DE Comparison 10v10

Here we'll repeat the above, but for a DE comparison 10 versus 10 samples
and 1500 DE genes are added. This will be done for 100 simulations.

## Generate a list of SB results objects

```{r, results='hide', message=FALSE}
sampleSize <- 10
nDE <- 1500

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de10, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))
} else {
  de10 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))
}
```

## Plot average results over replications

Plot results.

```{r}
res10d <- plotsim_standardize(de10, alpha = seq(0.01, 0.20, 0.01))

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res10d, met="rejections",filter_set = excludeSet, 
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

plotEffectSize(de5, covname="log2FC")
plotEffectSize(de5, covname="log2FC", DE=FALSE)

plotEffectSize(de5, covname="ind_covariate")
plotEffectSize(de5, covname="ind_covariate", DE=FALSE)

if (numberMethodsReject(res10d, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(de10, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, "; skipping upset plot")
}
```
